Tools for identifying unexpectedly low microfossil count sums

### 2023-06-26 count_check_MS.qmd
moved otherwise cannot find targets cache

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
execute: 
  echo: false
  error: false
---

@@

# Packages loaded for convenience when knitting without running r_make
library("tidyverse")
#library("rjt.misc")

#packages only loaded in this file
library("kableExtra")
library("english")

@@

tar_load("summarised_bird_singletons")
tar_load("summarised_bird")
tar_load("testate_summ")
tar_load("last_chance_summ")
tar_load("pollen_summ")
tar_load("hershop_pollen_summ")
tar_load("wildhorse")
tar_load("pollen_summ1")
tar_load("pollen_summ2")
tar_load("pollen3")
tar_load("pollen1")
tar_load("pollen2")
tar_load("owens_summ1")
tar_load("owens_summ")
tar_load("owens_est_summ")
tar_load("diatom1_summ")
tar_load("diatom2_summ")
tar_load("chironomid1_summ")
tar_load("marine1_summ")
tar_load("chironomid2_summ")

@@

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r tar_read(pollen_download_date)`.

@@

All analyses were done in R v. `r getRversion()` [@R] and used the packages 
extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
numbers v. `r packageVersion("numbers")` [@numbers], 
countSum v. `r packageVersion("countSum")` [@countSum], 
neotoma v. `r packageVersion("neotoma")` [@neotoma], 
bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Packages drake v. `r packageVersion("drake")` [@drake], 
bookdown v. `r packageVersion("bookdown")` [@bookdown1] and
renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
Code to reproduce all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms) and [zenodo.org](zenodo.org).
All issues identified in the Neotoma database were reported to the maintainers.

@@

Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets |> format()` pollen datasets. 
Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
`r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some pollen datasets with unexpected properties are discussed in section @sec-unexpected-count-data.

```{r}
#| label: tbl-multitaxa
#| tbl-cap: Properties of different data sets.
tar_read(count_summary) |> 
  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
  kbl(x = ., booktabs = TRUE, escape = FALSE,
      col.names = linebreak(str_wrap(names(.), width = 10))
    ) |> 
  kable_styling(latex_options = c("basic", "hold_position"))
```

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: manuscript_extra/countMS2.bib
csl: manuscript_extra/elsevier-harvard_rjt.csl
execute: 
  echo: false
  error: false
---

@@

# Packages loading
library("targets")
library("tidyverse")
#library("rjt.misc")

library("gt")
library("english")

@@

# tar_load("summarised_bird_singletons")
# tar_load("summarised_bird")
# tar_load("testate_summ")
 tar_load("last_chance_summ")
# tar_load("pollen_assem_res")
# tar_load("pollen_dataset_res")
 tar_load("pollen_results")
 tar_load("hershop_pollen_summ")
 tar_load("wildhorse")
 tar_load("pollen3")
 tar_load("pollen1")
 tar_load("pollen2")
# tar_load("owens_summ1")
# tar_load("owens_summ")
# tar_load("owens_est_summ")
 tar_load("diatom1_summ")
 tar_load("diatom2_summ")
 tar_load("chironomid1_summ")
 tar_load("marine1_summ")
# tar_load("chironomid2_summ")

@@

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r targets::tar_meta(pollen_data) |> pull(time) |> format("%d %B %Y")`.

@@

All analyses were done in R v. `r getRversion()` [@R] and used the packages 
extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
numbers v. `r packageVersion("numbers")` [@numbers], 
countSum v. `r packageVersion("countSum")` [@countSum], 
neotoma2 v. `r packageVersion("neotoma2")` [@neotoma2], 
bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Packages targets v. `r packageVersion("targets")` [@targets], 
quarto v. `r packageVersion("quarto")` [@bookdown1] and
renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
Code to reproduce all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms) and [zenodo.org](zenodo.org).
All issues identified in the Neotoma database were reported to the maintainers.

@@

Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_results$n_datasets |> format()` pollen datasets. 
Half of the assemblages without singletons are in just `r pollen_results$half_no_sing` datasets.
`r Words(pollen_results$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_results$dataset_gcd1`% of the datasets have any such assemblages.
Some pollen datasets with unexpected properties are discussed in section @sec-unexpected-count-data.

```{r}
#| label: tbl-multitaxa
#| tbl-cap: Properties of different data sets.
tar_read(count_summary) |> 
  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
  gt(x = _)
# , booktabs = TRUE, escape = FALSE,
#       col.names = linebreak(str_wrap(names(.), width = 10))
#     ) |> 
#   kable_styling(latex_options = c("basic", "hold_position"))
```
